Comparing shared and distributed memory performance: A case study on Matrix Transposition and Symmetry Check\\

### 2025-01-12 latex/paper2.tex
second-assignment: contribution and methodology

Distributed-memory matrix transposition is a well discussed problem.
Choi, Dongarra and Walker \cite{b3} presented an implementation
based on the block cyclic data distribution. In their approach, the
matrix is divided into blocks of arbitrary size. If the numbers of rows and
columns are relatively prime, then the algorithm consists of a two-dimension
complete exchange of information where each processor sends a sub block
to every other processor. Binary exchange was shown to be the faster communication
scheme \cite{b4} which completes the exchange in $log_2 P$ steps where
$P$ is the number of processors. If the number of rows and columns
are not relatively prime, processors are grouped into a number of groups
equal to the greatest common divisor of the number of rows and columns, and
communication is overlapped between groups using
non-blocking point-to-point communication. Bowman and Roberts \cite{b5} later
presented an hybrid approach leveraging both shared and distributed memory.
First, block of the matrix are transposed locally. Eckstein and Matyasfalvi
\cite{b6} studied the distributed problem on wide or tall unstructured sparce matrices.
Their implementation assignes nonzero elements of the matrix evenly
between processes by partitioning the matrix by sections of non zero
elements of the same length. Juan Gómez-Luna et. al. \cite{b7} presented
several GPU applications and an heuristic to detect the optimal approach.

@@

TODO
In this section, implementations of both \textit{matTranspose}
and \textit{checkSymm} using MPI are presented.

@@

@@

\begin{thebibliography}{00}
\bibitem{b0} P. Grun, Introduction to InfiniBand for End Users, InfiniBand
Trade Assoc., Beaverton, OR, USA, 2010. [Online]. Available:
cw.infinibandta.org/document/dl/7268
\bibitem{b-1} Kalia, A., Kaminsky, M., and Andersen, D. G. (2016). Design Guidelines for High Performance RDMA Systems. 2016 USENIX Annual Technical Conference (USENIX ATC 16), 437–450. https://www.usenix.org/conference/atc16/technical-sessions/presentation/kalia 
\bibitem{b-2} Message Passing Interface Forum. (2023). MPI: A Message-Passing Interface Standard Version 4.1. Chapter 8.7 "Caching". https://www.mpi-forum.org/docs/mpi-4.1/mpi41-report.pdf 
\bibitem{b1} G. M. Shipman, T. S. Woodall, R. L. Graham, A. B. Maccabe and P. G. Bridges, "Infiniband scalability in Open MPI," Proceedings 20th IEEE International Parallel and Distributed Processing Symposium, Rhodes, Greece, 2006, pp. 10 pp.-, doi: 10.1109/IPDPS.2006.1639335.
\bibitem{b2} J. Liu et al., "Design and implementation of MPICH2 over InfiniBand with RDMA support," 18th International Parallel and Distributed Processing Symposium, 2004. Proceedings., Santa Fe, NM, USA, 2004, pp. 16-, doi: 10.1109/IPDPS.2004.1302922.
\bibitem{b3} Jaeyoung Choi, Jack J. Dongarra, David W. Walker.  Parallel Matrix Transpose Algorithms on Distributed Memory Concurrent Computers.  Mathematical Sciences Section Oak Ridge National Laboratory P.O. Box 2008, Bldg. 6012 Oak Ridge, TN 37831-6367
\bibitem{b4} S. Takkella and S. Seidel, "Complete exchange and broadcast algorithms for meshes," Proceedings of IEEE Scalable High Performance Computing Conference, Knoxville, TN, USA, 1994, pp. 422-428, doi: 10.1109/SHPCC.1994.296674.
\bibitem{b5} John C. Bowman and Malcolm Roberts. Adaptive Matrix Transpose Algorithms for Distributed Multicore Processors
\bibitem{b6} Eckstein, J., Matyasfalvi, G. (2018). Efficient Distributed-Memory Parallel Matrix-Vector Multiplication with Wide or Tall Unstructured Sparse Matrices. https://arxiv.org/abs/1812.00904 
\bibitem{b7} J. Gómez-Luna, I. -J. Sung, L. -W. Chang, J. M. González-Linares, N. Guil and W. -M. W. Hwu, "In-Place Matrix Transposition on GPUs," in IEEE Transactions on Parallel and Distributed Systems, vol. 27, no. 3, pp. 776-788, 1 March 2016, doi: 10.1109/TPDS.2015.2412549.
\end{thebibliography}
\vspace{12pt}

>>>>

Distributed-memory matrix transposition is a well discussed problem.
Choi, Dongarra and Walker \cite{b3} presented an implementation
based on the block cyclic data distribution. In their approach, the
matrix is divided into blocks of arbitrary size. If the numbers of rows and
columns are relatively prime, then the algorithm consists of a two-dimension
complete exchange of information where each processor sends a sub block
to every other processor. Binary exchange was shown to be the faster communication
scheme \cite{b4} which completes the exchange in $log_2 P$ steps where
$P$ is the number of processors. If the number of rows and columns
are not relatively prime, processors are grouped into a number of groups
equal to the greatest common divisor of the number of rows and columns, and
communication is overlapped between groups using
non-blocking point-to-point communication. Bowman and Roberts \cite{b5} later
presented an hybrid approach leveraging both shared and distributed memory.
Eckstein and Matyasfalvi
\cite{b6} studied the distributed problem on wide or tall unstructured sparce matrices.
Their implementation assignes nonzero elements of the matrix evenly
between processes by partitioning the matrix by sections of non zero
elements of the same length. Juan Gómez-Luna et. al. \cite{b7} presented
several GPU applications and an heuristic to detect the optimal approach.

@@

In this section, implementations of both \textit{matTranspose}
and \textit{checkSymm} using MPI are presented, starting with
\textit{matTranspose}. We benchmark the bandwidth and speedup
of different MPI implementations compared to a serial and to a OpenMPI one.
Different compiler optimizations will also be compared. All benchmarks
will be tested against every matrix size from $2^2$ to $2^{12}$,
we will assume the number of processes is divisible by the matrix size
and the square root is an integer number for ease of implementation,
however the algorithms can be generalised with few modifications.
The algorithms for \textit{matTranspose} will now be explained.

\textit{matTranspose} takes a $N \times N$ input square matrix
\textit{input}, a matrix of the same shape \textit{ouput}, and the
size of the two matrices \textit{size}. The analyzed algorithms are the following:
\begin{itemize}
\item \textit{matTranspose}: naive implementation of matTranspose, iterating over values of the input matrix and storing them in the output matrix with indexes of rows and columns swapped.
\item \textit{matTransposeMPI}: this implementation takes full advantage
  of MPI datatypes to test bandwidth. MPI has the possibility to define a
  custom datatype to access memory, we used \textit{MPI\_Type\_contiguous}
  to define a row type and \textit{MPI\_Type\_vector} + \textit{MPI\_Type\_create\_resized}
  for a column type. The root process scatters rows with \textit{MPI\_Scatter} and
  retrieves columns with \textit{MPI\_Gather}, therefore
  transposing the matrix. Each process receives a block of rows and
  sends back the same block to the root but as column type. The algorithm
  automatically assign the correct number of rows/column to each process based on the
  number of processes so that the entire matrix is transposed.
  For example, in a $16x16$ matrix with $4$ processes, each processes
  receives $\frac{16}{4}=4$ rows and sends the same number back as
  columns.
  This implementation fully tests the communication
  bandwidth and latency by relying only on communication and MPI
  functionalities.
\item \textit{matTransposeMPIBlock}: this implementation divides the
  original matrix into a mesh of square sub-matrixes with side length equal to
  $\frac{N}{\sqrt{\#processes}}$ using MPI datatypes. Each sub-matrix is assigned
  to a different processes. The root process sends the sub-matrix to the
  other processes with \textit{MPI\_Scatterv} and receives the same
  number of sub-matrices but in transposed order with \textit{MPI\_Gatherv},
  for example process[0][1] will sand to root[1][0].
  Each process tranposes its sub-matrix and sends It back to the root.
  This approach tests a different communication method.
\end{itemize}
The algorithm for \textit{checkSymm} will now be explained. The algorithm
takes an $N \times N$ square matrix \textit{input}, a size \textit{size}
and returns true if the matrix is symmetrix, false otherwise.
The algorithm the following:
\begin{itemize}
\item \textit{checkSymm}: naive implementation of checkSymm. It Iterates over the upper half of the matrix and testing if every value is equal to the symmetric value.
\item \textit{checkSymmMPI}: similar to \textit{matTransposeMPIBlock},
but each process receives both a sub-matrix and the transposed sub-matrix,
that is, if we treated each sub-matrix as a single element of a matrix,
the element in the tranposed position will be received. Each process compares the two
sub-matrices to check if they are symmetric. The result is computed
via \textit{MPI\_Reduce}.
\end{itemize}

@@

Experiments were conducted on a cluster with 96 Intel(R) Xeon(R) Gold
6252N CPUs (2.30GHz) featuring 32K L1d and L1i caches, 1022K L2 cache,
36608K L3 cache (64-bit cache lines), 1 thread per core, and 4GB RAM,
running Linux 3.10. The algorithms were benchmarked using \textit{valFuzz},
with benchmarks compiled using gcc-9.1.0. Four benchmark sets were
compiled with different optimization levels: none, -O1, -O2 and "-O3
-Ofast -march=native".

To ensure statistical significance, each benchmark was executed 100
times, recording wall-clock execution time statistics: minimum, maximum,
median, mean, standard deviation, Q1, and Q3. Cache interference was
minimized by clearing it before each run by writing to an array twice
the L3 cache size. Input matrices, consistent across algorithms, were
initialized with uniformly random floats (-1 to 1) generated via a
Linear Congruential Generator (LCG) \cite{b8} with parameters a=1664525,
c=1013904223, m=232 \cite{b9}, and seed 1337.

To expedite random number generation, values were precomputed at
compile time using \textit{tenno-tl}. Additional \textit{tenno-tl}
features, such as ranges and arrays, were also employed.

@@

\begin{thebibliography}{00}
\bibitem{b0} P. Grun, Introduction to InfiniBand for End Users, InfiniBand
Trade Assoc., Beaverton, OR, USA, 2010. [Online]. Available:
cw.infinibandta.org/document/dl/7268
\bibitem{b-1} Kalia, A., Kaminsky, M., and Andersen, D. G. (2016). Design Guidelines for High Performance RDMA Systems. 2016 USENIX Annual Technical Conference (USENIX ATC 16), 437–450. https://www.usenix.org/conference/atc16/technical-sessions/presentation/kalia 
\bibitem{b-2} Message Passing Interface Forum. (2023). MPI: A Message-Passing Interface Standard Version 4.1. Chapter 8.7 "Caching". https://www.mpi-forum.org/docs/mpi-4.1/mpi41-report.pdf 
\bibitem{b1} G. M. Shipman, T. S. Woodall, R. L. Graham, A. B. Maccabe and P. G. Bridges, "Infiniband scalability in Open MPI," Proceedings 20th IEEE International Parallel and Distributed Processing Symposium, Rhodes, Greece, 2006, pp. 10 pp.-, doi: 10.1109/IPDPS.2006.1639335.
\bibitem{b2} J. Liu et al., "Design and implementation of MPICH2 over InfiniBand with RDMA support," 18th International Parallel and Distributed Processing Symposium, 2004. Proceedings., Santa Fe, NM, USA, 2004, pp. 16-, doi: 10.1109/IPDPS.2004.1302922.
\bibitem{b3} Jaeyoung Choi, Jack J. Dongarra, David W. Walker.  Parallel Matrix Transpose Algorithms on Distributed Memory Concurrent Computers.  Mathematical Sciences Section Oak Ridge National Laboratory P.O. Box 2008, Bldg. 6012 Oak Ridge, TN 37831-6367
\bibitem{b4} S. Takkella and S. Seidel, "Complete exchange and broadcast algorithms for meshes," Proceedings of IEEE Scalable High Performance Computing Conference, Knoxville, TN, USA, 1994, pp. 422-428, doi: 10.1109/SHPCC.1994.296674.
\bibitem{b5} John C. Bowman and Malcolm Roberts. Adaptive Matrix Transpose Algorithms for Distributed Multicore Processors
\bibitem{b6} Eckstein, J., Matyasfalvi, G. (2018). Efficient Distributed-Memory Parallel Matrix-Vector Multiplication with Wide or Tall Unstructured Sparse Matrices. https://arxiv.org/abs/1812.00904 
\bibitem{b7} J. Gómez-Luna, I. -J. Sung, L. -W. Chang, J. M. González-Linares, N. Guil and W. -M. W. Hwu, "In-Place Matrix Transposition on GPUs," in IEEE Transactions on Parallel and Distributed Systems, vol. 27, no. 3, pp. 776-788, 1 March 2016, doi: 10.1109/TPDS.2015.2412549.
\bibitem{b8} Rotenberg, A. (1960). "A New Pseudo-Random Number Generator". Journal of the ACM. 7 (1): 75–77
\bibitem{b9} Numerical Recipes ranqd1, Chapter 7.1, An Even Quicker Generator, Eq. 7.1.6
parameters from Knuth and H. W. Lewis
\end{thebibliography}
\vspace{12pt}
